---
title: "Finding Plateaus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding Plateaus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
cpal <- paste0('#', c('003E51', '007DBA', 'FFFFFF', 'D6D2C4', 'B78B20'))
```

Along with this package, this vignette uses the `gratia` and `mgcv` packages for comparsons. I'll start with the first example from `mgcv`.

```{r setup, results='hold', fig.width=5}
# library(curvish)
library(mgcv)
library(gratia)
library(ggplot2)


#Below from ?gam
set.seed(2) ## simulate some data... 
d <- gamSim(1,n=200,dist="normal",scale=2)
fit <- gam(y~s(x0)+s(x1)+s(x2)+s(x3),data=d, method = 'REML')
summary(fit)
par(mar=c(1,1,1,1))
plot(fit,pages=1,residuals=TRUE)  ## show partial residuals
plot(fit,pages=1,seWithMean=TRUE) ## `with intercept' CIs
## run some basic model checks, including checking
## smoothing basis dimensions...
gam.check(fit)
```

# Derivatives

```{r}
eps = 1e-07
res = 50
fit <- gam(y~s(x2),data=d, method = 'REML')
deriv_su <- gratia::derivatives(fit, term = 'x2', n = res, eps = eps, level = .95, 
                                interval = 'simultaneous', unconditional = TRUE, n_sim = 10000)
deriv_sc <- gratia::derivatives(fit, term = 'x2', n = res, eps = eps, level = .95, 
                                interval = 'simultaneous', unconditional = FALSE, n_sim = 10000)
deriv_cu <- gratia::derivatives(fit, term = 'x2', n = res, eps = eps, level = .95, 
                                interval = 'confidence', unconditional = TRUE)
deriv_cc <- gratia::derivatives(fit, term = 'x2', n = res, eps = eps, level = .95, 
                                interval = 'confidence', unconditional = FALSE)
```

## Sources of uncertainty

I'll start off by considering the important issue of simultaneous confidence
intervals described in this [blog
post](https://fromthebottomoftheheap.net/2014/06/16/simultaneous-confidence-intervals-for-derivatives/)
by the author of the `gratia` package. Essentially, if we are to compare,
point by point, the confidence interval, we need to correct for multiple
comparisons. See below for how this changes the width of the interval around
the first derivative plot. Also note how making the interval unconditional on
the smooth terms slightly widens it. This is to add in the uncertainty about
the exact shape of the smooth.

The Bayesian method does not face this particular multiple comparison
problem. I use the posterior of the derivative to compute the posterior for
the difference between the derivative at a given age to the derivative at the
age where the median posterior derivative is at its maximum (i.e., at `r sprintf('[INSERT HERE]')`
years). I then compute the 95% credible interval for that statistic at each age.
If at each age 95% of the posterior falls in a certain range, then we can also be sure that
95% of the posterior across all ages falls in this range, so we are not inflating the probability
that we make a mistake in deciding that this range includes the true region of steepest slope.

In the plots below, you will notice that the more sources of uncertainty we account for
in the ML models (that is, the non-Bayesian models), the wider the confidence regions
are. The widest interval is when we account for the multiple tests (i.e., a test at each age
we're interested in examining the derivative), and when we account for uncertainty in the
exact shape of the spline. "Simul" refers to the use of simultaneous confidence intervals
that account for multiple testing, whereas "Conf" refers to traditional confidence intervals.
"Uncon" refers to intervals that are unconditional on the choice of shape of the spline and "Cond"
refers to intervals that are conditional on the exact trajectory as estimated.

The derivative of the Bayesian model falls somewhere in the middle, and has a slightly different shape.

```{r}
ggplot(deriv_su, aes(x = data, y = derivative)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper, color = 'Simul/Uncond', fill = 'Simul/Uncond'), alpha = .25) + 
  geom_ribbon(data = deriv_sc, aes(ymin = lower, ymax = upper, color = 'Simul/Cond', fill = 'Simul/Cond'), alpha = .25) +
  geom_ribbon(data = deriv_cu, aes(ymin = lower, ymax = upper, color = 'Conf/Uncond', fill = 'Conf/Uncond'), alpha = .25) +
  geom_ribbon(data = deriv_cc, aes(ymin = lower, ymax = upper, color = 'Conf/Cond', fill = 'Conf/Cond'), alpha = .25) +
  scale_color_manual(breaks = c('Simul/Uncond', 'Simul/Cond', 'Conf/Uncond', 'Conf/Cond', 'Bayes'),
                     values = c(cpal[c(1:2, 4:5)], '#000000'),
                     name = 'Type of interval') + 
  scale_fill_manual(breaks = c('Simul/Uncond', 'Simul/Cond', 'Conf/Uncond', 'Conf/Cond', 'Bayes'),
                     values = c(cpal[c(1:2, 4:5)], '#000000'),
                     name = 'Type of interval') + 
  labs(x = 'x2', y = 'y') + 
  geom_line() + 
  theme_minimal()
```



# Bayesian GAM and derivative

Using brms.

```{r}
fit_b <- brms::brm(brms::bf(y ~ s(x0) + s(x1) + s(x2) + s(x3)), 
                   data=d,
                   chains = 4, cores = 4,
                   iter = 4500, warmup = 2000,
                   control = list(adapt_delta = .999, max_treedepth = 20),
                   file = 'fit_b')
summary(fit_b)
plot(fit_b)
plot(brms::conditional_smooths(fit_b, smooths = 's(x2)'))

deriv_b <- curvish::derivatives(fit_b, term = 'x2', n = res, eps = eps)

ggplot(deriv_su, aes(x = data, y = derivative)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper, color = 'Simul/Uncond', fill = 'Simul/Uncond'), alpha = .15) + 
  geom_ribbon(data = deriv_sc, aes(ymin = lower, ymax = upper, color = 'Simul/Cond', fill = 'Simul/Cond'), alpha = .15) +
  geom_ribbon(data = deriv_cu, aes(ymin = lower, ymax = upper, color = 'Conf/Uncond', fill = 'Conf/Uncond'), alpha = .15) +
  geom_ribbon(data = deriv_cc, aes(ymin = lower, ymax = upper, color = 'Conf/Cond', fill = 'Conf/Cond'), alpha = .15) +
  geom_ribbon(data = p_deriv_sum, aes(x = x2, y = mean, ymin = l, ymax = u, color = 'Bayes', fill = 'Bayes'), 
              alpha = .7) + 
  geom_line(data = deriv_b, aes (x = x2, y = mean)) + 
  scale_color_manual(breaks = c('Simul/Uncond', 'Simul/Cond', 'Conf/Uncond', 'Conf/Cond', 'Bayes'),
                     values = c('#aaaaaa', cpal[c(1:2, 4:5)]),
                     name = 'Type of interval') + 
  scale_fill_manual(breaks = c('Simul/Uncond', 'Simul/Cond', 'Conf/Uncond', 'Conf/Cond', 'Bayes'),
                    values = c('#222222', cpal[c(1:2, 4:5)]),
                    name = 'Type of interval') + 
  labs(x = 'x2', y = 'y') + 
  theme_minimal()
```

